---
title: 'Forecasting of Brent Crude Oil Price: ARIMA Modelling'
author: Rami Kh
date: '2018-12-19'
slug: forecasting-of-brent-crude-oil-price-arima-modelling
categories: []
tags: []
---


##Time Series: Definition

<p style='text-align: justify;'>A sequence of of data points being recorded at specific times. </p>


###Example of Time Series: 

```{r}
plot(AirPassengers)
```


###Basic Details of Time Series:

```{r warning=FALSE, message=FALSE}
library(astsa)
```

```{r}
start(AirPassengers)
end(AirPassengers)
frequency(AirPassengers)
length(AirPassengers)
```

##The Components of Time Series:

**1- Trend:** <p style='text-align: justify;'>Overall upward or downward pattern of the data. This trend in the series might be a result of political, economic or technological factors among other factors.</p>

**2- Seasonality:** <p style='text-align: justify;'>Regular pattern of upward/downward fluctuations that occur within certain period of time due to external factors such as waether, customs, holidays ..etc. </p>

**3- Noise:** <p style='text-align: justify;'>Unsystemtaic fluctuations that happen due to random variation or unforseen events. </p>

##Decomposition of Time Series

```{r}
plot(decompose(AirPassengers))
```

##Basic Time Series Models

**1- White Noise (WN):** <p style='text-align: justify;'>The simplest exapmple of stationary process. It is a purely random process that has constant mean and variace and has zero autocovariance. Each observation in the white noise series is uncorrelated with all other values in the sequence. </p>

```{r}
WN <- arima.sim(model = list(order = c(0,0,0)), n = 200)
ts.plot(WN)
```


**2- Random Walk:** <p style='text-align: justify;'>A time series said to follow a random walk if the first differences are random. The random walk process can be deemed as the simplest example of non-stationary process. </p>

$$y_t = y_{t-1} + \varepsilon_t$$
where $\varepsilon_t$ is white noise. 

```{r}
RW <- arima.sim(model = list(order = c(0,1,0)), n = 200)
ts.plot(RW)
```

**3- Autoregressive Model(AR):**<p style='text-align: justify;'>A time series in which $y_t$ depends only on its past values. </p>

$$y_{t} = c + \phi_{1}y_{t-1} + \phi_{2}y_{t-2} + \dots + \phi_{p}y_{t-p} + \varepsilon_{t}$$
Where:

when $\phi_1$ = 0, $y_t$ is equivalent to a white noise.

when $\phi_1$ = 1 and $c\ne0$, $y_t$ is equivalent to a random walk.

when $\phi_1$ = 1 and $cne0$, $y_t$ is equivalent to a random walk with a drift. 

```{r}
AR <- arima.sim(model = list(ar = 0.5), n = 200)
ts.plot(AR)
```

**4- Simple Moving Average(MA):**<p style='text-align: justify;'>A time series in which $y_t$ depends only on the error terms including the current and the previous ones. </p>

$$y_{t} = c + \varepsilon_t + \theta_{1}\varepsilon_{t-1} + \theta_{2}\varepsilon_{t-2} + \dots + \theta_{q}\varepsilon_{t-q}$$

Where:

when $\theta_1$ = 0, $y_t$ is equivalent to a white noise.

when $\theta_1$ $\ne0$, $y_t$ depends on both $\varepsilon_{t}$ and $\varepsilon_{t-1}$.

```{r}
MA <- arima.sim(model = list(ma = 0.5), n = 200)
ts.plot(MA)
```

**5- ARMA Model:** <p style='text-align: justify;'>A time series that combines AR model and MA model. In other words, the ARMA model is just an autoregressive model with autocorrelated errors.</p>

$$y_t = µ + \phi_{1}y_{t-1} + ... + \phi_{p}y_{t-p} + \theta_{1}\varepsilon_{t-1} + .... + \theta_{q}\varepsilon_{t-q}$$

```{r}
ARMA <- arima.sim(model = list(ar = 0.75, ma = 0.3), n = 200)
ts.plot(ARMA)
```

**6- ARIMA Model:**<p style='text-align: justify;'>A time series that combines AR modle and MA model after integreating the differences needed for stationarity. </p>

$$y'_{t} = c + \phi_{1}y'_{t-1} + \cdots + \phi_{p}y'_{t-p}  + \theta_{1}\varepsilon_{t-1} + \cdots + \theta_{q}\varepsilon_{t-q} + \varepsilon_{t}$$

where:

$y'_{t}$ is the differenced series.

$p$ = order of AR part.

$d$ = degree of first differencing.

$q$ = order of MA part.


##The Modelling of ARMA Models: Box-Jenkins Method

<p style='text-align: justify;'>The most popular thechnique in modelling and forecasting time series is Box-Jenkis Method (named after George Box and Gwilym Jenkins who wrote a book on time series analysis in 1976). The method follows three steps: </p>

<p style='text-align: justify;'>**1- Identification:** Assessing whether the time series is stationary, and if not, we should know how many differences are required to make it stationary. Assume that we have a non-stationary process as follows:</p>

```{r}
A <- arima.sim(model = list(order = c(0,1,0)), n = 200)
ts.plot(A)
```

<p style='text-align: justify;'>If we want to transform a non-stationary process into a stationary one, we can apply diffrencing on the time series. </p>

```{r}
Stationary <- diff(A)
ts.plot(Stationary)
```

<p style='text-align: justify;'>Then we need to identify the parameters of ARMA model for our data. How?</p>

<p style='text-align: justify;'>There are two diagnostic plots that can be used to help choosing the $p$ and $q$ parameters of ARMA model. </p>

<p style='text-align: justify;'>**1- Autocorrelation Function (AFC)**: The plot summarizes the correlation of an observation with lag value. </p>

<p style='text-align: justify;'>**2- Partial Autocorrelation Function (PACF)**: The plot summarizes the correlation of an observation that is not accounted for by prior lagged observation. </p>

###Rules of Autocorrelations for Chosing the Order of Model: 

<p style='text-align: justify;'>- If ACF tails off and PACF cuts off at lag p, this should be a strong indictaor of possible AR(p) model. </p>

```{r results="hide"}
AR <- arima.sim(model = list(ar = 0.9), n = 200)
acf2(AR)
```

<p style='text-align: justify;'>- If ACF cuts off at laq q and PACF tails off, this should be a strong indictaor of possible MA(q) model. </p>

```{r results="hide"}
MA <- arima.sim(model = list(ma = -0.5), n = 200)
acf2(MA)
```


<p style='text-align: justify;'>- If both ACF and PACF are tailing off, this should be a strong indicator of ARMA(p,q) model. </p>

```{r results="hide"}
ARMA <- arima.sim(model = list(ar = 0.65, ma = -0.4), n  = 200)
acf2(ARMA)
```


**2- Estimation:** The estimation of the of the model parameters involves using numerical method to minimize the error terms. </p>

```{r}
x <- arima.sim(list(order = c(1, 0, 1), ar = 0.9, ma = -0.4), n = 200)
fit_x <- sarima(x, p = 1, d = 0, q = 1)
```


<p style='text-align: justify;'>**3- Diagnostic Checking:** Assessing whether the model overfits the data. This means that the model is more complex than it needs to be and captures radnom noise in the data.</p>

**AIC/BIC:** measures the errors and penalizes for adding parameters and help us choosing the suitable model. 

<center>Average(observed - predicted)^2 + k (p + q)</center>

Where: 

AIC has k = 2 and BIC has k = log(k)




