---
title: 'Forecasting of Brent Crude Oil Price: ARIMA Modelling '
author: Rami Kh
date: '2018-12-22'
slug: forecasting-of-brent-crude-oil-price-arima-modelling
categories:
  - Time Series Analysis
tags: []
---


##Introduction:

<p style='text-align: justify;'>Crude oil is almost the most important commodities in the world. It is widely considered to be the lifeblood of our modern life and the crucial driver behind the radical transformations in the global economy over the last decades. The crude oil and its products are almost used in every single industry ranging from aviation to petrochemical. The world currently produces approximately 93 million barrels per day. The market of crude is bigger than all raw metal markets combined at approximate value of 1.7 trillion dollar a year. Given this exceptional importance, the price of crude oil is one of the key determinant of the global economic activity, the international trade, the stock prices, and the macroeconomic policies adopted by the countries worldwide. It has therefore been a subject of a great deal of research. An enormous number of governmental and international agencies, consultancies, hedge funds and think-tanks are always busy designing and evaluating the best approaches to forecast the crude oil prices and to capture its volatile trends. In this project, I will build an econometric model in order to study the dynamics of monthly Brent oil prices for the period from 1987 to 2018 using the [Box-Jenkins method](https://ramikh.netlify.com/post/forecasting-of-brent-crude-oil-price-arima-modelling/). I will use several ARIMA models in order to assess their forecasting performances using the relevant statistical tests and techniques. The data for this project is obtained from the Energy Information Administration (EIA) for the spot price of Brent crude in Europe in American Dollar for the mentioned period. </p>

 
###Packages used in Modelling and Forecasting

```{r}
library(ggplot2)
library(readxl)
library(urca)
library(tseries)
library(astsa)
```
 

##Methodology


**1- Identification:**

<p style='text-align: justify;'>Following the Box-Jenkins method, the first step in building our econometric model should be started with the assessment of the stationarity of the time series under consideration. If the time series proves to be non-stationary, it should be differenced until it becomes stationary. </p>

<p style='text-align: justify;'>Let us first import our data. </p>

```{r}
brent <- read.csv("C:/Users/Rami/Desktop/brent.csv")
head(brent)
```

<p style='text-align: justify;'>Since we are going to conduct a time series analysis, we need to transform the dataframe into time series object. </p>

```{r}
brent_ts <- ts(brent[,2], start = c(1987,5), frequency = 12)
```

<p style='text-align: justify;'>Let us then plot our series in order to capture its main charachteristics. </p>

```{r}
ts.plot(brent_ts)
```

<p style='text-align: justify;'>As we can see, our time series is far from being stationary. It has a positive trend in general, however, it includes vloatile periods as well. The price of crude oil has been relatively stable during 1980s and 1990s before increasing drastically starting from 2000 onward. The price has then experienced a remarkable slumps in 2009 as a result of the financal crisis. It then took an upward trend before sliding agains in 2015 as a result of OPEC policy to restore its market share after shale oil glut. </p>


As we can see, our time series is far from being stationary. It has a positive trend in general, however, 